Live Functional Programming with Typed Holes

### 2018-04-16 calculus.tex
revisions through the first 2 paragraphs of 3.3

\clearpage
\newcommand{\calculusSec}{Hazelnut Live, Formally}
\section{\protect\calculusSec}
\label{sec:calculus}
\input{fig-syntax}
\input{fig-bidirectional-typing}

We will now make the intuitions developed in the previous section formally precise by specifying the \HazelnutLive core calculus and  its accompanying metatheory.

The syntax of the core calculus specified in Fig.~\ref{fig:hazelnut-live-syntax} consists of types and expressions with holes. 
We distinguish between \emph{external} expressions, $e$, and \emph{internal} expressions, $d$. 
External expressions correspond to programs as entered by the programmer 
(see Sec.~\ref{sec:intro}\todo{or maybe Sec. 2?}{} for discussion of manual, semi-automated and fully automated hole entry methods). 
Each well-typed external expression (as specified in Sec.~\ref{sec:external-statics} below) expands to a well-typed internal expression (see Sec.~\ref{sec:expansion}) before it is evaluated (see Sec.~\ref{sec:evaluation}). 
We take this approach, notably also taken in the specification of Standard ML by \citet{Harper00atype-theoretic}, because (1) the external language supports type inference and explicit type ascriptions, $\hexp : \htau$, but it is formally simpler to eliminate ascriptions and specify a type assignment system when defining the dynamic semantics; and 
(2) we need additional syntactic machinery during evaluation for tracking hole closures and dynamic type casts. 
This machinery is inserted by the expansion step, rather than entered explicitly by the programmer. 
In this regard, the internal language is analogous to the cast calculus in the gradually typed lambda calculus \cite{DBLP:conf/snapl/SiekVCB15,Siek06a}, though as we will see the \HazelnutLive internal language goes beyond the cast calculus in several respects. We have mechanized these formal developments using the Agda proof assistant \cite{norell:thesis,norell2009dependently} 
(see Sec.~\ref{sec:agda-mechanization}). We have also implemented \HazelnutLive in a manner that maintains a novel end-to-end well-definedness invariant (see Sec.~\ref{sec:implementation}).

@@

We start with the type system of the \HazelnutLive external language,
which closely follows the \Hazelnut type system \cite{popl-paper},
except that (1) holes in \HazelnutLive each have a \emph{unqiue name},
and (2) for brievity of exposition, we consider a more austere core
language here that lacks numbers and sum types, which we discuss
(along with other systematic extensions) in
Appendix~\ref{sec:extensions}.

@@

The primary benefit of specifying the \HazelnutLive external language 
bidirectionally is that the programmer need not annotate each hole with a type. 
%
An empty hole is
written $\hehole{u}$, where $u$ is a name for the hole, which we tacitly assume is unique.
%
%(Holes in \Hazelnut are \emph{unnamed}).
%
Rule {SEHole}\todo{rule name macros}{} specifies that an empty hole synthesizes hole type, written $\tehole$.
%
If an empty hole appears where an expression of some other type is
expected, e.g. under an explicit ascription (governed by Rule {SAsc})
or in the argument position of a function application (governed by
Rule {SAp}, discussed below), we apply the \emph{subsumption rule},
Rule {ASubsume}, which specifies that if an expression $e$ synthesizes
type $\htau$, then it may be checked against any \emph{consistent}
type, $\htau'$.

@@

Fig.~\ref{fig:tconsistent} specifies the type consistency relation, written $\tconsistent{\htau}{\htau'}$, which specifies that two types are consistent if they differ only up to type holes in corresponding positions.
%
The hole type is consistent with every type, and so, by the subsumption rule, expression holes may appear where an expression of any type is expected. The type consistency relation here coincides with the type consistency relation from gradual type theory by identifying the hole type with the unknown type~\cite{Siek06a}.
%
Type consistency is reflexive and symmetric, but it is \emph{not} transitive.
%
This relation stands in contrast to subtyping, which is anti-symmetric and transitive; subtyping may be integrated into a gradual type system following \citet{Siek:2007qy}).

Non-empty expression holes, written $\hhole{\hexp}{u}$, behave similarly to empty holes.
%
Rule {SNEHole} specifies that a non-empty expression hole also synthesizes hole type as long as the expression inside the hole, $\hexp$, synthesizes some (arbitrary) type.
%
Non-empty expression holes therefore internalize the ``red squiggles'' that many editors display under or around type inconsistencies in a program.\todo{example?}\todo{briefly say something about binding inconsistencies?}

For the familiar forms of the lambda calculus, the rules again follow prior work.
%
For simplicity, the core calculus includes only a single base type~$b$ with a single constant~$c$, governed by Rule {SConst} (i.e. $b$ is the unit type).
%
%\matt{Extraneous and uninteresting:}
%By contrast, \Hazelnut instead defines a number type with a single operation, which we include in Appendix~\ref{sec:extensions} alongside various other standard extensions to the core calculus\todo{do this, say more?}. 
%
Rule {SVar} synthesizes the corresponding type from $\hGamma$.

For the sake of exposition, \HazelnutLive includes ``half-annotated'' lambdas, $\halam{x}{\htau}{\hexp}$, in addition to the unannotated lambdas, $\hlam{x}{\hexp}$, from \Hazelnut.
%
Half-annotated lambdas may appear in synthetic position according to Rule {SLam}, which is standard \cite{Chlipala:2005da}.
%
Unannotated lambdas may only appear where the expected type is known to be either an arrow type or the hole type, which is treated as if it were $\tarr{\tehole}{\tehole}$.
%
To avoid the need for two separate rules, Rule {ALam} uses the auxiliary relation $\arrmatch{\htau}{\tarr{\htau_1}{\htau_2}}$ in \Figref{fig:arrmatch}, which produces the matched arrow type $\tarr{\tehole}{\tehole}$ given the hole type, and operates as the identity on arrow types \cite{DBLP:conf/snapl/SiekVCB15,DBLP:conf/popl/GarciaC15}.
%
\Secref{sec:related} dicusses how \HazelnutLive might be enriched with
with ML-style type reconstruction~\cite{damas1982principal}, perhaps via
the approach outlined by~\citet{DBLP:conf/icfp/DunfieldK13}.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% To Cyrus from Matt:
%
% Why say the following here? --- It's discussing a different design that we didn't pursue here.  We should move such discussion to related work.
%
%Note that a system supporting ML-style type reconstruction \cite{damas1982principal} might include a synthetic rule for unannotated lambdas, e.g. as outlined by \citet{DBLP:conf/icfp/DunfieldK13}.
%%

The rule governing function application, Rule {SAp}, similarly treats an expression of hole type in function position as if it were of type $\tarr{\tehole}{\tehole}$ using the same matched arrow type judgement.

@@

Each well-typed external expression~$e$ expands to a well-typed internal expression~$d$, for evaluation.
%
\Figref{fig:expansion} defines this expansion, and \Figref{fig:hasType} defines type assignment for internal expressions.
%
\Secref{sec:evaluation} discusses internal expression evaluation.

As with the type system for the external language (above), 
we specify expansion bidirectionally \cite{DBLP:conf/ppdp/FerreiraP14}.
%
The synthetic expansion judgement~$\expandSyn{\hGamma}{\hexp}{\htau}{\dexp}{\Delta}$ synthesizes a type~$\htau$ from~$\hexp$.
%
Synthetic expansion produces an internal expression~$d$ and a hole context~$\hDelta$.
%
%We say more about hole contexts, which are used in the type assignment judgement, $\hasType{\Delta}{\hGamma}{d}{\htau}$, below.
We describe hole contexts, which serve as ``inputs'' to the type assignment judgement~$\hasType{\Delta}{\hGamma}{d}{\htau}$, further below. 
%
The analytic expansion judgement~$\expandAna{\hGamma}{\hexp}{\htau}{\dexp}{\htau'}{\Delta}$, checks~$\hexp$ against~$\htau$ and produces an expansion~$d$ of type~$\htau'$, and a hole context~$\hDelta$.
%
The following theorem establishes that expansions are well-typed and in the analytic case that the type $\htau'$ is consistent with $\htau$.
%
\begin{thm}[Typed Expansion]\label{thm:typed-expansion} ~
  \begin{enumerate}[nolistsep]
    \item
      If $\expandSyn{\hGamma}{\hexp}{\htau}{\dexp}{\Delta}$
      then $\hasType{\Delta}{\hGamma}{\dexp}{\htau}$.
    \item
      If $\expandAna{\hGamma}{\hexp}{\htau}{\dexp}{\htau'}{\Delta}$
      then $\tconsistent{\htau}{\htau'}$ and $\hasType{\Delta}{\hGamma}{\dexp}{\htau'}$.
  \end{enumerate}
\end{thm}
\noindent
%
%The reason analytic expansion produces an expansion of consistent
%type is because the subsumption rule, as previously discussed, allows
%us to check an external expression against any type consistent with
%the type the expression actually synthesizes, whereas every internal
%expression can be assigned at most one type, i.e. the following
%standard unicity property holds of the type assignment system.
%
Analytic expansion produces an expansion of consistent type:
%
Intuitively, the subsumption rule permits us
to check an external expression against any type consistent with the
type that the expression \emph{actually} synthesizes:
%
\begin{thm}[Type Assignment Unicity]
  If $\hasType{\Delta}{\hGamma}{\dexp}{\htau}$
  and $\hasType{\Delta}{\hGamma}{\dexp}{\htau'}$
  then $\htau=\htau'$.
\end{thm}
\noindent
Consequently, analytic expansion reports the type actually assigned to the expansion it produces.
%
For example, we can derive that $\expandAna{\hGamma}{c}{\tehole}{c}{b}{\emptyset}$.% where $\emptyset$ is the empty hole context.

Before describing the rules in detail, let us state a few other useful theorems. 
The following theorem establishes that an expansion exists for every well-typed external expression.
 \begin{thm}[Expandability] \label{thm:expandability}~
  \begin{enumerate}[nolistsep]
    \item
      If $\hsyn{\hGamma}{\hexp}{\htau}$
      then $\expandSyn{\hGamma}{\hexp}{\htau}{\dexp}{\Delta}$
      for some $\dexp$ and $\Delta$.
    \item
      If $\hana{\hGamma}{\hexp}{\htau}$
      then $\expandAna{\hGamma}{\hexp}{\htau}{\dexp}{\htau'}{\Delta}$
      for some $\dexp$ and $\htau'$ and $\Delta$.
  \end{enumerate}
\end{thm}
\noindent
The following theorem establishes that when an expansion exists, it is unique.
\begin{thm}[Expansion Unicity] \label{thm:expansion-unicity}~
  \begin{enumerate}[nolistsep]
    \item
      If $\expandSyn{\hGamma}{\hexp}{\htau}{\dexp}{\Delta}$
      and $\expandSyn{\hGamma}{\hexp}{\htau'}{\dexp'}{\Delta'}$
      then $\htau=\htau'$ and $\dexp=\dexp'$ and $\Delta=\Delta'$.
    \item
      If $\expandAna{\hGamma}{\hexp}{\htau_1}{\dexp}{\htau_2}{\Delta}$
      and $\expandAna{\hGamma}{\hexp}{\htau_1}{\dexp'}{\htau_2'}{\Delta'}$
      then $\dexp=\dexp'$ and $\htau_2=\htau_2'$ and $\Delta=\Delta'$.
  \end{enumerate}
\end{thm}
\noindent
The following theorem establishes that expansion generalizes external typing.\todo{rename correspondence to generality}
\begin{thm}[Expansion Generality] \label{thm:expansion-generality}~
  \begin{enumerate}[nolistsep]
    \item
      If $\expandSyn{\hGamma}{\hexp}{\htau}{\dexp}{\Delta}$
      then $\hsyn{\hGamma}{\hexp}{\htau}$.
    \item
      If $\expandAna{\hGamma}{\hexp}{\htau}{\dexp}{\htau'}{\Delta}$
      then $\hana{\hGamma}{\hexp}{\htau}$.
  \end{enumerate}
\end{thm}

The rules governing expansion of constants, variables and lambda expressions
--- Rules {ESConst}, {ESVar}, {ESLam} and {EALam}
--- and the corresponding type assignment rules
--- Rules {TAConst}, {TAVar} and {TALam}
--- mirror the typing rules from Fig.~\ref{fig:bidirectional-typing}.
%
Consequently, the corresponding cases of
Theorem~\ref{thm:typed-expansion}, Theorem~\ref{thm:expandability} and
Theorem~\ref{thm:expansion-generality} are straightforward.
%
To support type assignment, all lambdas in the internal language are
half-annotated---Rule {EALam} inserts the annotation when expanding an
unannotated external lambda based on the given type.
%
Some rules perform \emph{cast insertion}, making them more interesting
than the others; these rules include those for holes, and the those for
function applications and ascriptions.  We consider each below.

\subsubsection{Hole Expansion}\label{sec:hole-expansion} 
Rules {ESEHole}, {ESNEHole}, {EAEHole} and {EANEHole} govern the expansion of empty and non-empty expression holes to empty and non-empty \emph{hole closures}, $\dehole{u}{\sigma}{}$ and $\dhole{\dexp}{u}{\sigma}{}$, respectively. 
%
The hole name~$u$ on a hole closure identifies the external hole to which the hole closure corresponds.
%
%Note that
While we assume each hole name to be unique in the external language, due to substitution, evaluated (internal) terms may consist of multiple hole closures with the same name.
%
For example, the result from Fig.~\ref{fig:grades-example} shows four closures for the hole named 1\todo{update with actual number (four?) and hole name later}.
%
There, we number each hole closure for a given hole sequentially, \li{1:1}, \li{1:2} and so on, but this is strictly for the sake of presentation, so we omit hole closure numbers from the core calculus.

The hole expansion rules introduce hypotheses of the form~$\Dbinding{u}{\hGamma}{\htau}$ into the hole context~$\Delta$.
%
For each hole name~$u$, the hole context records a type~$\tau$, and a typing context~$\Gamma$.\footnote{
We use a hole context, rather than recording the typing context and type directly on each hole closure, to ensure that all closures for a hole name have the same typing context and type.}
%
We borrow this hole context notation from contextual modal type theory (CMTT) \cite{Nanevski2008}, identifying hole names with metavariables and hole contexts with modal contexts (we say more about the connection with CMTT below).
%
Each hole expansion rule records the ``current'' typing context under which the hole is expanded.
%
In the synthetic hole expansion rules~{ESEHole} and~{ESNEHole}, the generated hole context assigns the hole type~$\tehole$ to hole name~$u$, as in the typing rules.
%
However, the first two premises of the expansion subsumption rule~{EASubsume} disallow the use of subsumption for holes in analytic position.
%
Instead, we employ separate analytic rules~{EAEHole} and~{EANEHole}, which each record the checked type~$\tau$ in the hole context.
%
Consequently, we can use type assignment for the internal language --- the type assignment rules TAEHole and TANEHole in Fig.~\ref{fig:hasType} assign a hole closure for hole name~$u$ the corresponding type from the hole context.

Each hole closure also has an associated environment~$\sigma$ which consists of a finite substitution of the form $[d_1/x_1, ~\cdots, d_n/x_n]$ for $n \geq 0$.
%
The closure environment keeps a record of the substitutions that occur around the hole as evaluation occurs.
%
Initially, when no evaluation has yet occurred, the hole expansion rules generate an initial environment that consists of the identity substitution for the typing context associated with hole name~$u$ in hole context~$\Delta$, which we notate $\idof{\hGamma}$, and define as follows.
%
\begin{defn}[Identity Substitution] $\idof{x_1 : \tau_1, ~\cdots, x_n : \tau_n} = [x_1/x_1, ~\cdots, x_n/x_n]$
\end{defn}
\noindent
The type assignment rules for hole closures~{TAEHole} and {TANEHole} each require that the hole closure environment~$\sigma$ be consistent with the corresponding typing context, written as $\hasType{\Delta}{\hGamma}{\sigma}{\hGamma'}$.
%
Formally, we define this relation as follows:\todo{check that the definition in the Agda corresponds}
\begin{defn}[Substitution Typing]
$\hasType{\Delta}{\hGamma}{\sigma}{\hGamma'}$ iff $\domof{\sigma} = \domof{\hGamma'}$ and for each $x : \htau \in \hGamma'$ we have that $d/x \in \sigma$ and $\hasType{\Delta}{\hGamma}{d}{\tau}$.
\end{defn}
\noindent
It is easy to verify that the identity substitution satisfies this requirement, i.e. that $\hasType{\Delta}{\hGamma}{\idof{\hGamma}}{\hGamma}$.

Empty hole closures, $\dehole{u}{\sigma}{}$,  correspond to the metavariable closures (a.k.a. deferred substitutions) from CMTT, $\cmttclo{u}{\sigma}$.\todo{change notation from CMTT}{}
%
\Secref{sec:evaluation} defines how these closure environments evolve during evaluation.
%
Non-empty hole closures~$\dhole{d}{u}{\sigma}{}$ have no direct correspondence with a notion from CMTT (see Sec.~\ref{sec:resumption}).

>>>>

\clearpage
\newcommand{\calculusSec}{Hazelnut Live, Formally}
\section{\protect\calculusSec}
\label{sec:calculus}

% \def \TirNameStyle \Vtexttt{#1}{{#1}}

\input{fig-syntax}
\input{fig-bidirectional-typing}

We will now make the intuitions developed in the previous section formally precise by specifying a core calculus, which we call \HazelnutLive, and its accompanying metatheory.

\parahead{Overview} The syntax of the core calculus given in Fig.~\ref{fig:hazelnut-live-syntax} consists of types and expressions with holes. 
We distinguish between {external} expressions, $e$, and {internal} expressions, $d$. 
External expressions correspond to programs as entered by the programmer 
(see Sec.~\ref{sec:intro} for discussion of implicit, manual, semi-automated and fully automated hole entry methods). 
Each well-typed external expression (as specified in Sec.~\ref{sec:external-statics} below) expands to a well-typed internal expression (see Sec.~\ref{sec:expansion}) before it is evaluated (see Sec.~\ref{sec:evaluation}). 
We take this approach, notably also taken in the ``redefinition'' of Standard ML by \citet{Harper00atype-theoretic}, because (1) the external language supports type inference and explicit type ascriptions, $\hexp : \htau$, but it is formally simpler to eliminate ascriptions and specify a type assignment system when defining the dynamic semantics; and 
(2) we need additional syntactic machinery during evaluation for tracking hole closures and dynamic type casts. 
This machinery is inserted by the expansion step, rather than entered explicitly by the programmer. 
In this regard, the internal language is analogous to the cast calculus in the gradually typed lambda calculus \cite{DBLP:conf/snapl/SiekVCB15,Siek06a}, though as we will see the \HazelnutLive internal language goes beyond the cast calculus in several respects. We have mechanized these formal developments using the Agda proof assistant \cite{norell:thesis,norell2009dependently} 
(see Sec.~\ref{sec:agda-mechanization}). Rule names, written \rulename{SVar}, correspond to the names used in the mechanization. We have also implemented \HazelnutLive in a manner that maintains a novel end-to-end well-definedness invariant (see Sec.~\ref{sec:implementation}).

@@

We start with the type system of the \HazelnutLive external language,
which closely follows the \Hazelnut type system \cite{popl-paper}; we summarize the minor differences as they come up.

% except that (1) holes in \HazelnutLive each have a \emph{unique name};
% (2) for brevity of exposition, we removed numbers in favor of a simpler base type, $b$, with one value, $c$ (i.e. $b$ is the unit type); and 
% (3) we include both unannotated lambdas, $\hlam{x}{\hexp}$, and half-annotated lambdas, $\halam{x}{\htau}{\hexp}$.  which we discuss
% (along with other systematic extensions) in
% Appendix~\ref{sec:extensions}.

@@

The primary benefit of specifying the \HazelnutLive external language 
bidirectionally is that the programmer need not annotate each hole with a type. 
%
An empty hole is
written simply $\hehole{u}$, where $u$ is the hole name, which we tacitly assume is unique 
(holes in \Hazelnut were not named).
Rule \rulename{SEHole} specifies that an empty hole synthesizes hole type, written $\tehole$.
%
If an empty hole appears where an expression of some other type is
expected, e.g. under an explicit ascription (governed by Rule \rulename{SAsc})
or in the argument position of a function application (governed by
Rule \rulename{SAp}, discussed below), we apply the \emph{subsumption rule},
Rule \rulename{ASubsume}, which specifies that if an expression $e$ synthesizes
type $\htau$, then it may be checked against any \emph{consistent}
type, $\htau'$.

@@

Fig.~\ref{fig:tconsistent} specifies the type consistency relation, written $\tconsistent{\htau}{\htau'}$, which specifies that two types are consistent if they differ only up to type holes in corresponding positions.
%
The hole type is consistent with every type, and so, by the subsumption rule, expression holes may appear where an expression of any type is expected. The type consistency relation here coincides with the type consistency relation from gradual type theory by identifying the hole type with the unknown type~\cite{Siek06a}.
%
Type consistency is reflexive and symmetric, but it is \emph{not} transitive.
%
This stands in contrast to subtyping, which is anti-symmetric and transitive; subtyping may be integrated into a gradual type system following \citet{Siek:2007qy}.

Non-empty expression holes, written $\hhole{\hexp}{u}$, behave similarly to empty holes.
%
Rule \rulename{SNEHole} specifies that a non-empty expression hole also synthesizes hole type as long as the expression inside the hole, $\hexp$, synthesizes some (arbitrary) type.
%
Non-empty expression holes therefore internalize the ``red underline'' that many editors display under or around type inconsistencies in a program.

For the familiar forms of the lambda calculus, the rules again follow prior work.
%
For simplicity, the core calculus includes only a single base type~$b$ with a single constant~$c$, governed by Rule \rulename{SConst} (i.e. $b$ is the unit type).
%
%\matt{Extraneous and uninteresting:}
By contrast, \Hazelnut instead defines a number type with a single operation, which we include in Appendix~\ref{sec:extensions} alongside various other simple extensions to the calculus\todo{do this, say which ones}. 
%

Rule \rulename{SVar} synthesizes the corresponding type from $\hGamma$. 
For the sake of exposition, \HazelnutLive includes ``half-annotated'' lambdas, $\halam{x}{\htau}{\hexp}$, in addition to the unannotated lambdas, $\hlam{x}{\hexp}$, from \Hazelnut.
%
Half-annotated lambdas may appear in synthetic position according to Rule \rulename{SLam}, which is standard \cite{Chlipala:2005da}.
%
Unannotated lambdas may only appear where the expected type is known to be either an arrow type or the hole type, which is treated as if it were $\tarr{\tehole}{\tehole}$.
%
To avoid the need for two separate rules, Rule \rulename{ALam} uses the matching relation $\arrmatch{\htau}{\tarr{\htau_1}{\htau_2}}$ defined in \Figref{fig:arrmatch}, which produces the matched arrow type $\tarr{\tehole}{\tehole}$ given the hole type, and operates as the identity on arrow types \cite{DBLP:conf/snapl/SiekVCB15,DBLP:conf/popl/GarciaC15}.\footnote{A system supporting ML-style type reconstruction \cite{damas1982principal} might also include a synthetic rule for unannotated lambdas, e.g. as outlined by \citet{DBLP:conf/icfp/DunfieldK13}, but we stick to this simpler ``Scala-style'' local type inference scheme in this paper \cite{Pierce:2000ve,Odersky:2001lb}.}
%
% \Secref{sec:related} dicusses how \HazelnutLive might be enriched with
% with ML-style type reconstruction~\cite{damas1982principal}, perhaps via
% the approach outlined by~\citet{DBLP:conf/icfp/DunfieldK13}.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% To Cyrus from Matt:
%
% Why say the following here? --- It's discussing a different design that we didn't pursue here.  We should move such discussion to related work.
%
%Note that 
%%

The rule governing function application, Rule \rulename{SAp}, similarly treats an expression of hole type in function position as if it were of type $\tarr{\tehole}{\tehole}$ using the same matched arrow type judgement.

@@

Each well-typed external expression~$e$ expands to a well-typed internal expression~$d$, for evaluation.
%
\Figref{fig:expansion} specifies expansion, and \Figref{fig:hasType} specifies type assignment for internal expressions.
%
% \Secref{sec:evaluation} discusses internal expression evaluation.

As with the type system for the external language (above), 
we specify expansion bidirectionally \cite{DBLP:conf/ppdp/FerreiraP14}.
%
The synthetic expansion judgement~$\expandSyn{\hGamma}{\hexp}{\htau}{\dexp}{\Delta}$  
%
produces an expansion~$d$ and a hole context~$\hDelta$ when synthesizing type $\htau$ for $\hexp$. 
%
%We say more about hole contexts, which are used in the type assignment judgement, $\hasType{\Delta}{\hGamma}{d}{\htau}$, below.
We describe hole contexts, which serve as ``inputs'' to the type assignment judgement~$\hasType{\Delta}{\hGamma}{d}{\htau}$, further below. 
%
The analytic expansion judgement~$\expandAna{\hGamma}{\hexp}{\htau}{\dexp}{\htau'}{\Delta}$, produces an expansion~$d$ of type~$\htau'$, and a hole context~$\hDelta$, when checking~$\hexp$ against~$\htau$.
%
The following theorem establishes that expansions are well-typed and in the analytic case that the assigned type, $\htau'$, is consistent with provided type, $\htau$.
%
\begin{thm}[Typed Expansion]\label{thm:typed-expansion} ~
  \begin{enumerate}[nolistsep]
    \item
      If $\expandSyn{\hGamma}{\hexp}{\htau}{\dexp}{\Delta}$
      then $\hasType{\Delta}{\hGamma}{\dexp}{\htau}$.
    \item
      If $\expandAna{\hGamma}{\hexp}{\htau}{\dexp}{\htau'}{\Delta}$
      then $\tconsistent{\htau}{\htau'}$ and $\hasType{\Delta}{\hGamma}{\dexp}{\htau'}$.
  \end{enumerate}
\end{thm}
\noindent
%
%The reason analytic expansion produces an expansion of consistent
%type is because the subsumption rule, as previously discussed, allows
%us to check an external expression against any type consistent with
%the type the expression actually synthesizes, whereas every internal
%expression can be assigned at most one type, i.e. the following
%standard unicity property holds of the type assignment system.
%
The reason that $\htau'$ is only consistent with the provided type $\htau$ is because 
%
 the subsumption rule permits us
to check an external expression against any type consistent with the
type that the expression \emph{actually} synthesizes, whereas every internal
expression can be assigned at most one type, i.e. the following
standard unicity property holds of the type assignment system.
%
\begin{thm}[Type Assignment Unicity]
  If $\hasType{\Delta}{\hGamma}{\dexp}{\htau}$
  and $\hasType{\Delta}{\hGamma}{\dexp}{\htau'}$
  then $\htau=\htau'$.
\end{thm}
\noindent
Consequently, analytic expansion reports the type actually assigned to the expansion it produces.
%
For example, we can derive that $\expandAna{\hGamma}{c}{\tehole}{c}{b}{\emptyset}$.% where $\emptyset$ is the empty hole context.

Before describing the rules in detail, let us state two other guiding theorems. 
The following theorem establishes that an expansion exists for every well-typed external expression. 
The mechanization also establishes when an expansion exists, it is unique (not shown).
 \begin{thm}[Expandability] \label{thm:expandability}~
  \begin{enumerate}[nolistsep]
    \item
      If $\hsyn{\hGamma}{\hexp}{\htau}$
      then $\expandSyn{\hGamma}{\hexp}{\htau}{\dexp}{\Delta}$
      for some $\dexp$ and $\Delta$.
    \item
      If $\hana{\hGamma}{\hexp}{\htau}$
      then $\expandAna{\hGamma}{\hexp}{\htau}{\dexp}{\htau'}{\Delta}$
      for some $\dexp$ and $\htau'$ and $\Delta$.
  \end{enumerate}
\end{thm}
% \noindent
% The following theorem establishes that when an expansion exists, it is unique.
% \begin{thm}[Expansion Unicity] \label{thm:expansion-unicity}~
%   % \begin{enumerate}[nolistsep]
%   %   \item
%       If $\expandSyn{\hGamma}{\hexp}{\htau}{\dexp}{\Delta}$
%       and $\expandSyn{\hGamma}{\hexp}{\htau'}{\dexp'}{\Delta'}$
%       then $\htau=\htau'$ and $\dexp=\dexp'$ and $\Delta=\Delta'$.
%   %   \item
%   %     If $\expandAna{\hGamma}{\hexp}{\htau_1}{\dexp}{\htau_2}{\Delta}$
%   %     and $\expandAna{\hGamma}{\hexp}{\htau_1}{\dexp'}{\htau_2'}{\Delta'}$
%   %     then $\dexp=\dexp'$ and $\htau_2=\htau_2'$ and $\Delta=\Delta'$.
%   % \end{enumerate}
% \end{thm}
% \noindent
The following theorem establishes that expansion generalizes external typing.
\begin{thm}[Expansion Generality] \label{thm:expansion-generality}~
  \begin{enumerate}[nolistsep]
    \item
      If $\expandSyn{\hGamma}{\hexp}{\htau}{\dexp}{\Delta}$
      then $\hsyn{\hGamma}{\hexp}{\htau}$.
    \item
      If $\expandAna{\hGamma}{\hexp}{\htau}{\dexp}{\htau'}{\Delta}$
      then $\hana{\hGamma}{\hexp}{\htau}$.
  \end{enumerate}
\end{thm}

The rules governing expansion of constants, variables and lambda expressions
--- Rules \rulename{ESConst}, \rulename{ESVar}, \rulename{ESLam} and \rulename{EALam}
--- mirror the corresponding type assignment rules
--- Rules \rulename{TAConst}, \rulename{TAVar} and \rulename{TALam}
--- mirror the bidirectional typing rules from Fig.~\ref{fig:bidirectional-typing}.
%
% Consequently, the corresponding cases of
% Theorem~\ref{thm:typed-expansion}, Theorem~\ref{thm:expandability} and
% Theorem~\ref{thm:expansion-generality} are straightforward.
%
To support type assignment, all lambdas in the internal language are
half-annotated---Rule \rulename{EALam} inserts the annotation when expanding an
unannotated external lambda based on the given type.
%
The rules governing hole expansion, and the rules that perform \emph{cast insertion}---those governing
function application and type ascription---are more interesting. We consider each of these two groups of rules, in turn, below.

\subsubsection{Hole Expansion}\label{sec:hole-expansion} 
Rules \rulename{ESEHole}, \rulename{ESNEHole}, \rulename{EAEHole} and \rulename{EANEHole} govern the expansion of empty and non-empty expression holes to empty and non-empty \emph{hole closures}, $\dehole{u}{\sigma}{}$ and $\dhole{\dexp}{u}{\sigma}{}$. 
%
The hole name~$u$ on a hole closure identifies the external hole to which the hole closure corresponds.
%
While we assume each hole name to be unique in the external language, once evaluation begins, there may be multiple hole closures with the same name due to substitution.
%
For example, the result from Fig.~\ref{fig:grades-example} shows four closures for the hole named 1\todo{update with actual number (four?) and hole name later}{}.
%
There, we number each hole closure for a given hole sequentially, \li{1:1}, \li{1:2} and so on, but this is strictly for the sake of presentation, so we omit hole closure numbers from the core calculus.

%
For each hole, $u$, in an external expression, the hole context generated by expansion, $\Delta$, contains a hypothesis of the form~$\Dbinding{u}{\hGamma}{\htau}$, which records the hole's type, $\tau$, and the typing context, $\Gamma$, from where it appears in the original expression.\footnote{
We use a hole context, rather than recording the typing context and type directly on each hole closure, to ensure that all closures for a hole name have the same typing context and type.}
%
We borrow this hole context notation from contextual modal type theory (CMTT) \cite{Nanevski2008}, identifying hole names with metavariables and hole contexts with modal contexts (we say more about the connection with CMTT below).
%
% Each hole expansion rule records the ``current'' typing context under which the hole is expanded.
%
In the synthetic hole expansion rules~\rulename{ESEHole} and~\rulename{ESNEHole}, the generated hole context assigns the hole type~$\tehole$ to hole name~$u$, as in the external typing rules.
%
However, the first two premises of the expansion subsumption rule~\rulename{EASubsume} disallow the use of subsumption for holes in analytic position.
%
Instead, we employ separate analytic rules~\rulename{EAEHole} and~\rulename{EANEHole}, which each record the checked type~$\tau$ in the hole context.
%
Consequently, we can use type assignment for the internal language --- the type assignment rules \rulename{TAEHole} and \rulename{TANEHole} in Fig.~\ref{fig:hasType} assign a hole closure for hole name~$u$ the corresponding type from the hole context.

Each hole closure also has an associated environment~$\sigma$ which consists of a finite substitution of the form $[d_1/x_1, ~\cdots, d_n/x_n]$ for $n \geq 0$.
%
The closure environment keeps a record of the substitutions that occur around the hole as evaluation occurs.
%
Initially, when no evaluation has yet occurred, the hole expansion rules generate the identity substitution for the typing context associated with hole name~$u$ in hole context~$\Delta$, which we notate $\idof{\hGamma}$, and define as follows.
%
\begin{defn}[Identity Substitution] $\idof{x_1 : \tau_1, ~\cdots, x_n : \tau_n} = [x_1/x_1, ~\cdots, x_n/x_n]$
\end{defn}
\noindent
The type assignment rules for hole closures~{TAEHole} and {TANEHole} each require that the hole closure environment~$\sigma$ be consistent with the corresponding typing context, written as $\hasType{\Delta}{\hGamma}{\sigma}{\hGamma'}$.
%
Formally, we define this relation in terms of type assignment as follows:
\begin{defn}[Substitution Typing]
$\hasType{\Delta}{\hGamma}{\sigma}{\hGamma'}$ iff $\domof{\sigma} = \domof{\hGamma'}$ and for each $x : \htau \in \hGamma'$ we have that $d/x \in \sigma$ and $\hasType{\Delta}{\hGamma}{d}{\tau}$.
\end{defn}
\noindent
It is easy to verify that the identity substitution satisfies this requirement, i.e. that $\hasType{\Delta}{\hGamma}{\idof{\hGamma}}{\hGamma}$.

Empty hole closures, $\dehole{u}{\sigma}{}$,  correspond to the metavariable closures (a.k.a. deferred substitutions) from CMTT, $\cmttclo{u}{\sigma}$.
%
\Secref{sec:evaluation} defines how these closure environments evolve during evaluation.
%
Non-empty hole closures~$\dhole{d}{u}{\sigma}{}$ have no direct correspondence with a notion from CMTT (see Sec.~\ref{sec:resumption}).
